---
title: "RayBiotech DED-Cytokine Array Analysis "
subtitle: "e1-359"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(here)
library(kableExtra)
library(RColorBrewer)
```



# Tidying the Data

**Sample Guide:**
```{r echo=FALSE, warning=FALSE}
sample_names <- c("healthy-ser-cntrl", "CMM2-1b", "CMM2-1c", "83b", "76-OD", "81-OS", "83a", "73a-OD")

sample_codes <- NULL
for (i in 1:8) {
    code <- paste0("SA", i)
    sample_codes <- append(sample_codes, code)
}

sample_codes_names <- data_frame(sample_codes, sample_names)
sample_codes_names %>% 
  kable(align = "cc") %>% 
  kableExtra::kable_styling()
```

Some notes on the samples and the data:

  - The serum samples were collected by differnt individuals and may have been processed differently. For example, different protocols for isolating plasma or serum.
    - SA1 provided by the Knight BioLibrary
    - SA2 processed by Tammie Haley
    - SA3 processed by Tammie Haley
    - SA4 (83b) was collected by Paul Yang

## Convert to Long format

Load in output from RayBiotech Excel analysis tool
```{r}
#cytokine concentrations calculated from linear regression standard curves
cytokines_linReg <- read_excel(path = here("data/Stdev-sheet_unprotected.xlsx"), sheet = "Stdev_linear-reg")

#cytokine concentrations calculated from log-log regression standard curves
cytokines_logReg <- read_excel(path = here("data/Stdev-sheet_unprotected.xlsx"), sheet = "Stdev_log-reg")
```


Make data *long* rather than *wide* with `pivot_longer()`
```{r}
cytokines_linReg <-  pivot_longer(cytokines_linReg, cols = 2:ncol(cytokines_linReg), names_to = "sample", values_to = "concentration")
cytokines_logReg <- pivot_longer(cytokines_logReg, cols = 2:ncol(cytokines_logReg), names_to = "sample", values_to = "concentration")


cytokine_data_list <- list("cytokines_linReg" = cytokines_linReg)
                           #, "cytokines_logReg" = cytokines_logReg)
```


## Separate the concentration and SD into separate variables

Strategy = Use Regex to:

 - identify numbers *before* a "+" --> cytokine concentration value
 - identify numbers *after* a "-" --> standard deviation value
 
```{r}
#regex patterns to find numbers before and after the plus and minus symbols in the dataframe
before_plus_pattern <- "[:digit:]+"
after_minus_pattern <- "[:digit:]+$"

#pattern to recognize "+/-" and split string on either side
split_pattern <- "\\+/-"


#function for make new cols for SD and concentration using str_extract
tidy_split <- function(dataframe, pattern1, pattern2) {
  dataframe <- dataframe %>% 
    mutate(SD = str_extract(string = dataframe$concentration, pattern = pattern1),
            concentration = str_extract(string = dataframe$concentration, pattern = pattern2)) %>%
    mutate(SD = as.integer(SD),
           concentration = as.integer(concentration)) %>% 
    arrange(sample)
}

cytokine_data_list <- lapply(cytokine_data_list, tidy_split, pattern1 = after_minus_pattern, pattern2 = before_plus_pattern)

```
# Parametric Tests

Determine what kind of stats will be used for testing this data

```{r}
car::qqPlot(cytokine_data_list[[1]]$concentration)
```
Data do not appear parametric by visualization

```{r}
shapiro.test(cytokine_data_list[[1]]$concentration)
```

p << 0.05 : Data are not normally distributed

# Displaying the Data


## Heat Map {.tabset}
Given the size and complexity of the data, heatmap may be a good start to understand broad patterns. After exploring large-scale patterns, I can create other plots with more nuanced comparisons


Plotting function
```{r}
#remove SDs, convert to matrix, heatmap()
heatmap_plot3 <- function(dataframe,legend_name, plotvar = "concentration", multiplier = 1, ...) { 

  fill_value <- NULL
  if(multiplier == 1){
    fill_value <- unlist(dataframe[plotvar])
    legend_name_addon <- ""
  }
  else if (multiplier == "log2"){
    fill_value <- unlist(log2(dataframe[plotvar]))
    legend_name_addon <- "(Log2)"
  }
 
  dataframe %>%
    select(!"SD") %>%
      ggplot(mapping = aes(x = cytokine, y = sample, fill = fill_value ))+
      geom_tile() +
      xlab("Cytokine") +
      ylab("Sample") +
      scale_fill_continuous(type = "viridis", name = paste(legend_name, " ", legend_name_addon))+
      theme_classic()+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
}

```

### Serum and Aqueous Humor Together {.tabset}

These data were normalized to positive controls on the array, but not to any of the positive control *samples* we included (such as healthy control serum or aqueous humor)

#### Raw Concentration Values

```{r, fig.width=12, fig.height=5}
lapply(cytokine_data_list, heatmap_plot3 , legend_name = "Concentration", multiplier = "1")
```


#### Log2 Transformation

The data span a large range of concentrations so I'm plotting it on a log scale for legibility
```{r, fig.width=12, fig.height=5}
lapply(cytokine_data_list, heatmap_plot3 , legend_name = "Concentration", multiplier = "log2")
```


### Serum Only Data {.tabset}

Since we have a healthy serum positive control (*SA1*), I can normalize the MAR serum to this sample:

  - Fold Change = y/x
  - Log2 FC = log2(y) - log2(x)

```{r, fig.width=12, fig.height=5}
#keep only the serum samples in each list (SA1 - SA4)
serum_samples <- c("SA1", "SA2", "SA3","SA4")
cytokine_serum_list <- lapply(cytokine_data_list, filter, sample == "SA2" | sample == "SA1" | sample == "SA3" | sample == "SA4")

#function for normalizing to SA1; "Fold Change" (y/x)
SA1_normalize <- function(df){
  df <- df %>% 
    mutate(FC = (concentration/concentration[df$sample == "SA1"]),
           FC = round(FC, digits = 3)) %>% 
    mutate(log2_FC = log2(concentration) - log2(concentration[df$sample == "SA1"]),
           log2_FC = round(log2_FC, digits = 3))
}


#list with dataframes that only have serum samples and concentration is normalized to SA1
cytokine_serum_list <- lapply(cytokine_serum_list, SA1_normalize)

```



#### Fold Change in Concentration

Fold Change = y/x
```{r, fig.width=12, fig.height=5}
lapply(cytokine_serum_list, heatmap_plot3, plotvar = "FC", legend_name = "Fold Change", multiplier = 1)
```

#### Log2 Fold Change in Concentration

Log2 FC = log2(y) - log2(x)
```{r, fig.width=12, fig.height=5}
lapply(cytokine_serum_list, heatmap_plot3, plotvar = "log2_FC", legend_name = "Fold Change", multiplier = 1)
```

**Observations**

  - IL-4 is not present in SA1 (concentration == 0), so there are several *infinite* or *NA* values in the normalized data that come from dividing by 0


## Volcano Plot

After calculating p-values I can identify the cytokines undergoing the most significant with a volcano plot

**Statistical** Test: `Wilcox Test`

  - Variables: 1
  - Quantitative data
  - Non-parametric
  
Determined using the [flow chart for statistical tests](https://statsandr.com/blog/files/overview-statistical-tests-statsandr.pdf)
